---
title: "Data preprocessing.Rmd"
author: "Capucine LE CAM LIGIER"
date: "2024-04-01"
format: 
  html:
    code-fold: true
---

# I. Load Required Packages

```{r global-options, include=FALSE}
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE, echo=FALSE)
```

```{r }
packages <- c("dada2", "phyloseq", "readr", "Biostrings", "DECIPHER", "ape", "ShortRead", "tidyverse", "vegan", "microbiome", "BiocManager", "ggplot2", "dplyr", "tidyr", "knitr", "rmarkdown")
lapply(packages, library, character.only = TRUE)
```

# II. Working Directory and Reference Databases

Define the folder where reference databases (SILVA) are stored:

```{r}
refdb_folder <- here::here("data", "refdb")
if (!dir.exists(refdb_folder)) dir.create(refdb_folder, recursive = TRUE)
```

Download SILVA reference databases if not already available:

```{r}
options(timeout = 1200)

silva_train_set <- file.path(refdb_folder, "silva_nr99_v138.1_train_set.fa.gz")
silva_species_assignment <- file.path(refdb_folder, "silva_species_assignment_v138.1.fa.gz")

if (!file.exists(silva_train_set)) {
  download.file("https://zenodo.org/record/4587955/files/silva_nr99_v138.1_train_set.fa.gz", silva_train_set, quiet = TRUE)
}

if (!file.exists(silva_species_assignment)) {
  download.file("https://zenodo.org/record/4587955/files/silva_species_assignment_v138.1.fa.gz", silva_species_assignment, quiet = TRUE)
}
```

# III. Input Files

```{r}
devtools::load_all()
```

Define paths to raw sequencing data:

```{r}
path_to_fastqs <- here::here("data", "raw")
```

List forward and reverse read files:

```{r}
fnFs <- sort(list.files(path_to_fastqs, pattern = "_1.fastq.gz", full.names = TRUE))
fnRs <- sort(list.files(path_to_fastqs, pattern = "_2.fastq.gz", full.names = TRUE))
```

Extract sample names from file names:

```{r}
sample_names <- basename(fnFs) |> strsplit(split = "_") |> sapply(head, 1)
```

# IV. Quality Control

Generate quality profile plots:

```{r}
quality_folder <- here::here("outputs", "dada2", "quality_plots")
if (!dir.exists(quality_folder)) dir.create(quality_folder, recursive = TRUE)

qualityprofile(fnFs, fnRs, file.path(quality_folder, "quality_plots.pdf"))
```

# V. Primer Removal

Create output directory and define primer sequences (V3--V4 region):

```{r}
path_to_trimmed_reads <- here::here("outputs", "dada2", "trimmed")
if (!dir.exists(path_to_trimmed_reads)) dir.create(path_to_trimmed_reads, recursive = TRUE)

primer_fwd  <- "CCTACGGGNGGCWGCAG"
primer_rev  <- "GACTACHVGGGTATCTAATCC"
```

Trim primers using cutadapt:

```{r}
(primer_log <- primer_trim(
  forward_files = fnFs,
  reverse_files = fnRs,
  primer_fwd = primer_fwd,
  primer_rev = primer_rev,
  output_dir = path_to_trimmed_reads,
  min_size = 200
))
```

# VI. Filtering and Trimming

Define output paths and filenames:

```{r}
path_to_filtered_reads <- here::here("outputs", "dada2", "filtered")
if (!dir.exists(path_to_filtered_reads)) dir.create(path_to_filtered_reads, recursive = TRUE)

nopFw <- sort(list.files(path_to_trimmed_reads, pattern = "_1", full.names = TRUE))
nopRv <- sort(list.files(path_to_trimmed_reads, pattern = "_2", full.names = TRUE))

filtFs <- file.path(path_to_filtered_reads, basename(fnFs))
filtRs <- file.path(path_to_filtered_reads, basename(fnRs))
names(filtFs) <- sample_names
names(filtRs) <- sample_names
```

Run filterAndTrim():

```{r}
(out <- dada2::filterAndTrim(
  fwd = nopFw,
  filt = filtFs,
  rev = nopRv,
  filt.rev = filtRs,
  minLen = 150,
  matchIDs = TRUE,
  maxN = 0,
  maxEE = c(1, 2),
  truncQ = 2
))
```

# VII. Denoising with DADA2

## 1. Error Learning

```{r}
errF <- dada2::learnErrors(filtFs, randomize = TRUE, multithread = TRUE)
errR <- dada2::learnErrors(filtRs, randomize = TRUE, multithread = TRUE)
```

Visualize error models:

```{r}
dada2::plotErrors(errF, nominalQ = TRUE)
dada2::plotErrors(errR, nominalQ = TRUE)
```

## 2. Dereplication and Denoising

```{r}
derepFs <- dada2::derepFastq(filtFs, verbose = TRUE)
derepRs <- dada2::derepFastq(filtRs, verbose = TRUE)

dadaFs <- dada2::dada(derepFs, err = errF, multithread = TRUE)
dadaRs <- dada2::dada(derepRs, err = errR, multithread = TRUE)
```

## 3. Merging Paired Reads

```{r}
mergers <- dada2::mergePairs(
  dadaF = dadaFs,
  derepF = derepFs,
  dadaR = dadaRs,
  derepR = derepRs,
  maxMismatch = 0,
  verbose = TRUE
)
```

## 4. ASV Table and Chimera Removal

```{r}
seqtab <- dada2::makeSequenceTable(mergers)
seqtab_nochim <- dada2::removeBimeraDenovo(seqtab, method = "consensus", multithread = TRUE, verbose = TRUE)
```

# VIII. Taxonomic Assignment

```{r}
taxonomy <- dada2::assignTaxonomy(
  seqs = seqtab_nochim,
  refFasta = silva_train_set,
  taxLevels = c("Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Species"),
  multithread = TRUE,
  minBoot = 60
)

taxonomy <- dada2::addSpecies(
  taxonomy,
  silva_species_assignment,
  allowMultiple = FALSE
)
```

# IX. Export Outputs

## 1. Save Objects

```{r}
export_folder <- here::here("outputs", "dada2", "asv_table")
if (!dir.exists(export_folder)) dir.create(export_folder, recursive = TRUE)

saveRDS(seqtab_nochim, file = file.path(export_folder, "seqtab_nochim.rds"))
saveRDS(taxonomy, file = file.path(export_folder, "taxonomy.rds"))
```

## 2. ASV IDs and Tables

```{r}
asv_seq <- colnames(seqtab_nochim)
ndigits <- nchar(length(asv_seq))
asv_id <- sprintf(paste0("ASV_%0", ndigits, "d"), seq_along(asv_seq))
row.names(taxonomy) <- colnames(seqtab_nochim) <- names(asv_seq) <- asv_id

taxonomy_export <- df_export(taxonomy, new_rn = "asv")
seqtab_nochim_export <- t(seqtab_nochim) |> df_export(new_rn = "asv")
```

## 3. Write to Files

```{r}
write.table(taxonomy_export, file = file.path(export_folder, "taxonomy.tsv"), quote = FALSE, sep = "\t", row.names = FALSE)
write.table(seqtab_nochim_export, file = file.path(export_folder, "asv_table.tsv"), quote = FALSE, sep = "\t", row.names = FALSE)

cat(paste0(">", names(asv_seq), "\n", asv_seq), sep = "\n", file = file.path(export_folder, "asv.fasta"))
```

## 4. Log Table

```{r}
getN <- function(x) sum(dada2::getUniques(x))

log_table <- data.frame(
  input = primer_log$in_reads,
  with_fwd_primer = primer_log$`w/adapters`,
  with_rev_primer = primer_log$`w/adapters2`,
  with_both_primers = out[, 1],
  filtered = out[, 2],
  denoisedF = sapply(dadaFs, getN),
  denoisedR = sapply(dadaRs, getN),
  merged = sapply(mergers, getN),
  nonchim = rowSums(seqtab_nochim),
  perc_retained = rowSums(seqtab_nochim) / out[, 1] * 100
)
rownames(log_table) <- sample_names
```

```{r}
df_export(log_table, new_rn = "sample") |>
  write.table(file = file.path(export_folder, "log_table.tsv"), quote = FALSE, sep = "\t", row.names = FALSE)
```

# X. Create Phyloseq Object

```{r}
input_dir <- here::here("outputs", "dada2", "asv_table")

asv_table <- read.table(file = file.path(input_dir, "asv_table.tsv"), header = TRUE, sep = "\t", row.names = 1)
asv_table <- subset(asv_table, select = -X2S337)

taxonomy <- read.table(file = file.path(input_dir, "taxonomy.tsv"), header = TRUE, sep = "\t", row.names = 1)

asv_seq <- Biostrings::readDNAStringSet(filepath = file.path(input_dir, "asv.fasta"), format = "fasta")

file_path <- here::here("data", "context", "contexte.txt")
context <- read.table(file = file_path, header = TRUE, sep = "\t")
rownames(context) <- context[, 1]
context <- context[, -1]
```

## Assemble Phyloseq Object

```{r}
physeq <- phyloseq::phyloseq(
  phyloseq::otu_table(asv_table, taxa_are_rows = TRUE),
  phyloseq::tax_table(as.matrix(taxonomy)),
  phyloseq::sample_data(context),
  phyloseq::refseq(asv_seq)
)
```

## Remove Mitochondria and Chloroplasts

```{r}
physeq <- subset_taxa(physeq, Family != "Mitochondria")
physeq <- subset_taxa(physeq, Order != "Chloroplast")
physeq <- subset_taxa(physeq, Phylum != "NA")
```

## Export

```{r}
export_folder <- here::here("data")
saveRDS(object = physeq,
        file = file.path(export_folder, "physeq.rds"))
```
