---
title: "Data analysis : alpha and beta diversity.Rmd"
author: "Capucine LE CAM LIGIER"
date: "2024-04-01"
format: 
  html:
    code-fold: true
---

# I. Load Required Packages

```{r global-options, include=FALSE}
# Global chunk settings
knitr::opts_chunk$set(message=FALSE, warning=FALSE, error=FALSE, echo=FALSE)
```

```{r }
# Load required libraries for microbiome analysis and plotting
packages <- c(
  "dada2", "phyloseq", "readr", "Biostrings", "DECIPHER", "ape", "ShortRead", "tidyverse",
  "vegan", "microbiome", "BiocManager", "ggplot2", "dplyr", "tidyr", "knitr",
  "rmarkdown", "readxl", "patchwork", "scales", "here", "viridis",
  "ggpubr", "rstatix")
lapply(packages, library, character.only = TRUE)
```

# II. Load Data

```{r}
# Load pre-processed phyloseq object
physeq <- readRDS(here("data", "physeq.rds"))
```

#III. Alpha diversity

## 1. Rarefaction, filtering, and taxonomic aggregation

```{r}
# Rarefy the dataset to an even sequencing depth
physeq_rar <- rarefy_even_depth(physeq, sample.size = 13000)
# Transform to relative abundance
physeq_rar_relab <- transform_sample_counts(physeq_rar, function(x) x / sum(x))
# Filter rare taxa (< 0.1%)
physeq_rar_filter <- filter_taxa(physeq_rar_relab, function(x) sum(x) >= 0.001, TRUE)
# Aggregate taxa to Family level
physeq_fam <- tax_glom(physeq_rar_filter, taxrank = "Family", NArm = FALSE)
```

## 2. Compute and merge alpha diversity indices

```{r}
# Calculate alpha diversity indices
# Calculate alpha diversity indices
richness_data <- plot_richness(
  physeq_rar,
  measures = c("Observed", "Shannon", "Simpson")
)$data

df_alpha <- as.data.frame(richness_data) %>%
  filter(variable %in% c("Observed", "Shannon", "Simpson")) %>%
  drop_na(type, value) %>%
  mutate(type = factor(type, levels = c("gut", "liver", "skin")))
```

## 3. Statistical analyses

```{r}
anova_results <- df_alpha %>%
  group_by(variable) %>%
  anova_test(value ~ type)

tukey_results <- df_alpha %>%
  group_by(variable) %>%
  tukey_hsd(value ~ type)

stat_labels <- tukey_results %>%
  mutate(
    group1 = factor(group1, levels = c("gut", "liver", "skin")),
    group2 = factor(group2, levels = c("gut", "liver", "skin"))
  ) %>%
  group_by(variable) %>%
  arrange(group1, group2, .by_group = TRUE) %>%
  mutate(
    y.position = c(
      max(df_alpha$value[df_alpha$variable == unique(variable)], na.rm = TRUE) * 1.05,
      max(df_alpha$value[df_alpha$variable == unique(variable)], na.rm = TRUE) * 1.10,
      max(df_alpha$value[df_alpha$variable == unique(variable)], na.rm = TRUE) * 1.15
    )[seq_len(n())]
  ) %>%
  ungroup()

anova_results
tukey_results
```

## 4. Fig 1. Alpha diversity across tissue types

```{r}
family_colors <- scales::hue_pal()(length(unique(df_alpha$family)))

Fig1 <- ggplot(df_alpha, aes(x = type, y = value, fill = family, color = family)) +
  geom_boxplot(alpha = 0.8) +
  facet_wrap(~ variable, scales = "free") +
  stat_pvalue_manual(
    stat_labels,
    label = "p.adj.signif",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.005,
    inherit.aes = FALSE,
    hide.ns = FALSE
  ) +
  scale_x_discrete(limits = c("gut", "liver", "skin")) +
  scale_color_manual(values = family_colors) +
  scale_fill_manual(values = family_colors) +
  labs(x = NULL, y = NULL, fill = "Fish family", color = "Fish family") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, size = 16),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(size = 14, face = "italic"),
    strip.text = element_text(size = 14, face = "bold"),
    strip.background = element_rect(fill = "white", color = "black")
  )

Fig1
```

```{r}
#Save
dir.create(here("outputs"), showWarnings = FALSE)
ggsave(
  filename = here("outputs", "Fig1.tiff"),plot = Fig1,dpi = 300,width = 10,height = 9,units = "in",device = "tiff")
```

# IV. Barplot by family

```{r}
# Keep most abundant families
physeq_fam <- prune_taxa(taxa_sums(physeq_fam) > 0.5, physeq_fam)

# Melt to long format for plotting
ps_data <- psmelt(physeq_fam)

# Define color palette
family_colors <- c("brown","black", "magenta", "green", "darkblue", "red", "orange", "pink", "yellow", "grey", "blue", "lightgreen", "cyan")

# Plot relative abundances per class
Fig2 <- ggplot(ps_data, aes(x = Class, y = Abundance, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~type, ncol = 3, scales = "free_y") +
  scale_fill_manual(name = "Family", values = family_colors) +
  labs(x = "Class", y = "Relative Abundance (%)") +
  guides(fill = guide_legend(ncol = 1)) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, face = "italic", size = 14),
    axis.text.y = element_text(size = 12),
    axis.title = element_text(size = 16),
    legend.title = element_text(size = 16, face = "bold"),
    legend.text = element_text(face = "italic", size = 14),
    strip.text = element_text(face = "bold", size = 14),
    strip.background = element_rect(fill = "white", color = "black")
  )

ggsave(filename = here::here("outputs", "Fig2.tiff"), plot = Fig2, dpi = 300, width = 10, height = 8, units = "in", device = "tiff")
```

# V. Barplot by genus per fish species

```{r}
# Aggregate at Genus level and filter
physeq_gen <- tax_glom(physeq_rar_filter, taxrank = "Genus")
physeq_gen <- prune_taxa(taxa_sums(physeq_gen) > 0.5, physeq_gen)
genus_data <- psmelt(physeq_gen)

# Set Genus color palette
palette_family <- c(
  "Acinetobacter" = "#FFFACD", "Akkermansia" = "#FFDFBA", "Aliivibrio" = "red",
  "Bifidobacterium" = "#BAFFC9", "Candidatus Hepatoplasma" = "#BAE1FF", "Cutibacterium" = "magenta",
  "Enterovibrio" = "orange", "Escherichia-Shigella" = "darkblue", "Lactobacillus" = "grey76",
  "Ligilactobacillus" = "brown", "Limosilactobacillus" = "#FFC3A0", "Mycoplasma" = "purple",
  "Photobacterium" = "#4cfaea", "Pseudoalteromonas" = "#b4f561", "Pseudomonas" = "yellow",
  "Psychrobacter" = "pink", "Vibrio" = "#3149fc", "Weissella" = "#ffe9cc")

# Relabel fish species
genus_data$specie <- factor(
  genus_data$specie,
  levels = c(
    "Argyropelecus_hemigymnus", "Argyropelecus_olfersii", "Benthosema_glaciale",
    "Symbolophorus_veranyi", "Notoscopelus_kroyeri", "Lobianchia_gemellarii",
    "Lampanyctus_crocodilus", "Electrona_risso", "Myctophum_punctatum",
    "Chauliodus_sloani", "Stomias_boa", "Malacosteus_niger",
    "Melanocetus_johnsonii", "Sigmops_elongatus", "Eurypharynx_pelecanoides",
    "Nansenia_oblita", "Serrivomer_beanii"),
  labels = c(
    "Argyropelecus hemigymnus", "Argyropelecus olfersii", "Benthosema glaciale",
    "Symbolophorus veranyi", "Notoscopelus kroyeri", "Lobianchia gemellarii",
    "Lampanyctus crocodilus", "Electrona risso", "Myctophum punctatum",
    "Chauliodus sloani", "Stomias boa", "Malacosteus niger",
    "Melanocetus johnsonii", "Sigmops elongatus", "Eurypharynx pelecanoides",
    "Nansenia oblita", "Serrivomer beanii"))
```


```{r}
# Plot genus composition by species
Fig3 <- ggplot(genus_data, aes(x = specie, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", position = "fill") +
  facet_wrap(~type, scales = "free_x") +
  scale_fill_manual(values = palette_family) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = "Fish species", y = "Relative abundance", fill = "Genus") +
  theme_bw() +
    theme(
    axis.text.x = element_text(angle = 90, hjust = 1, face = "italic", size = 11),
    axis.text.y = element_text(hjust = 1, size = 12),
    axis.title.x = element_text(size = 16),
    axis.title.y = element_text(size = 16),
    legend.title = element_text(size = 14, face = "bold"),
    legend.text = element_text(face = "italic", size = 14),
    strip.text = element_text(face = "bold", size = 14),
    strip.background = element_rect(fill = "white", color = "black")
  )
```

```{r}
ggsave(filename = here::here("outputs", "Fig3.tiff"), plot = Fig3, dpi = 300, width = 10, height = 7, units = "in", device = "tiff")
```

# VII. PCoA Ordination (Gut Only)

```{r}
GUT <- subset_samples(physeq_fam, type == "gut")
Fam_gut <- prune_taxa(taxa_sums(GUT) > 0.3, GUT)
Fam_gut <- subset_samples(Fam_gut, !is.na(sample_data(Fam_gut)$migrant_pattern))
sample_data(Fam_gut)$migrant_pattern <- factor(as.character(sample_data(Fam_gut)$migrant_pattern))
matrix_ordination_gut_BC <- ordinate(Fam_gut, "MDS", distance = "bray")
```

```{r}
# Samples ordination
fish_gut <- phyloseq::plot_ordination(Fam_gut, matrix_ordination_gut_BC, 
                                      type = "samples", color = "family") +  
  geom_point(size = 8, alpha = 1) +  
  theme_bw() +  
  scale_color_viridis(
    discrete = TRUE,
    option = "magma",
    name = "Family" 
  ) +
  guides(
    color = guide_legend(title = "Fish families")  
  ) +
  coord_fixed() +
  theme(
    axis.title.x = element_text(size = 22, face = "bold"),  
    axis.title.y = element_text(size = 22, face = "bold"),  
    axis.text.x  = element_text(size = 22),                 
    axis.text.y  = element_text(size = 22),                 
    legend.title = element_text(size = 24, face = "bold"),  
    legend.text  = element_text(size = 22, face = "italic"),
    legend.position = "right",
    legend.justification = c(0, 1)
    )
# Taxa ordination
taxa_gut <- phyloseq::plot_ordination(Fam_gut, matrix_ordination_gut_BC, 
                                      color = "Family", type = "taxa") +
  geom_point(size = 8, alpha = 1) + 
  theme_bw() +  
  guides(
    color = guide_legend(title = "Taxa")
  ) +
  coord_fixed() +
  theme(
    axis.title.x = element_text(size = 22, face = "bold"),  
    axis.title.y = element_text(size = 22, face = "bold"),  
    axis.text.x  = element_text(size = 22),                 
    axis.text.y  = element_text(size = 22),                 
    legend.title = element_text(size = 24, face = "bold"),  
    legend.text  = element_text(size = 22, face = "italic"),
    legend.position = "right",
    legend.justification = c(0, 1)
  )
# Trait ordination
trait_gut <- phyloseq::plot_ordination(
  Fam_gut, matrix_ordination_gut_BC,
  color = "trophic_level", type = "samples",
  shape = "migrant_pattern"
) +
  geom_point(size = 8, alpha = 1) +
  theme_bw() +
  scale_color_viridis_c(option = "viridis", name = "Trophic Position") +
  labs(shape = "Migration Pattern") +  
  coord_fixed() +
  theme(
    axis.title.x = element_text(size = 22, face = "bold"),  
    axis.title.y = element_text(size = 22, face = "bold"),  
    axis.text.x  = element_text(size = 22),                 
    axis.text.y  = element_text(size = 22),                 
    legend.title = element_text(size = 24, face = "bold"),  
    legend.text  = element_text(size = 22),
    legend.position = "right",
    legend.justification = c(0, 1))
```

```{r}
# Combined ordination figure
Fig4 <- trait_gut / fish_gut / taxa_gut +
  plot_layout(ncol = 1) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 30, face = "bold")  )

ggsave(filename = here::here("outputs", "Fig4.tiff"), plot = Fig4, dpi = 300, width = 13, height =20, units = "in", device = "tiff")
```

#VIII. Bioluminescent genera analysis

## 1. Extract target genera

```{r}
# Define the list of genera known to contain bioluminescent species
target_genera <- c("Vibrio", "Photobacterium", "Aliivibrio", "Enterovibrio")

# Extract taxonomic and abundance data from the phyloseq object
tax_data <- as.data.frame(tax_table(physeq))
otu_data <- as.data.frame(otu_table(physeq))

# Identify ASVs that belong to the target genera
target_asv <- rownames(tax_data)[tax_data$Genus %in% target_genera]

# Subset OTU table for selected ASVs only
otu_filtered <- otu_data[target_asv, ]

# Transform the OTU table to long format for plotting
otu_long <- otu_filtered %>%
  rownames_to_column(var = "ASV") %>%
  pivot_longer(-ASV, names_to = "ID", values_to = "Abundance")

# Re-add taxonomic annotations
tax_data <- tax_data %>% rownames_to_column(var = "ASV")
otu_long <- left_join(otu_long, tax_data, by = "ASV")

# Add sample metadata to the long-format OTU table
metadata <- as.data.frame(sample_data(physeq))
otu_long <- left_join(otu_long, metadata, by = "ID")
```

## 2. Plot Bioluminescent Genera Distribution

```{r}
# Filter data for gut samples and remove any with missing migration info
otu_gut <- otu_long %>%
  filter(type == "gut", !is.na(migrant_pattern))

# Create stacked barplot showing the distribution of bioluminescent genera
Fig5 <- ggplot(otu_gut, aes(x = family, y = Abundance, fill = Genus)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(diet_guild ~ migrant_pattern, scales = "free_y") +
  scale_fill_manual(
    values = c(
      "Vibrio" = "blue",
      "Photobacterium" = "turquoise",
      "Aliivibrio" = "red",
      "Enterovibrio" = "orange"
    )
  ) +
  labs(
    x = "Fish family",
    y = "Abundance (read counts)",
    fill = "Genera"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    axis.title = element_text(face = "bold", size = 14),
    axis.text.x = element_text(angle = 90, hjust = 1, size = 14, face = "italic"),
    axis.text.y = element_text(size = 14),
    legend.title = element_text(face = "bold", size = 14),
    legend.text = element_text(face = "italic", size = 14)
  )
```

```{r}
# Save the figure in the output directory using here::here()
ggsave(filename = here::here("outputs", "Fig5.tiff"), plot = Fig5, dpi = 300, width = 10, height = 10, units = "in", device = "tiff")
```
